---
title: "01_data-cleaning"
author: "David Henderson"
date: "2018-08-30"
output: html_document
---

## Introduction

This is the code used to import and clean the provided data files for the Hackathon.


###Packages

Load packages required...

```{r}
library(tidyverse)
library(magrittr) #Pipe operators
library(sf)    #maps and shapefile reading
theme_set(theme_minimal(base_size = 16))
```

###Data

Load data - 

Note that the synthetic data is for the use of the workshop only and is therefore included in the `.gitignore` file. This means data is on the local computer only and is not pushed to the GitHub repository.

I've downloaded some population estimates for each of the provinces which will help with some of the analysis from the Statistics Canada website [here](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000901).

I've also downloaded a shapefile of Canadian provinces from [this link](https://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/bound-limit-2011-eng.cfm) as I think it will be nice to map some of the variables.

```{r load_data}
raw_data <- read_csv(here("raw_data/ipdln_synth_final.csv"))
shapefile <- read_sf(here("raw_data/wether_gov_shapefile/province.shp"))
```


##Clean data

###Glimpse

The first thing I'll do is get a quick skim of the `raw-data` file to get an idea of its structure

```{r glimpse}
glimpse(raw_data)
```

Ok. All these variables are categorical and I need to label the values as per the metadata held in the `metadata` file folder (see [the github repo](https://github.com/davidhen/ipdln_hackathon) for files). 

###Missing data

Before I factorise everything - I'd like to check if there is any missing data in the data frame

```{r visdat}
map_df(raw_data, function(x) sum(is.na(x)))
```

Great, there are no missing values.

###Clean names and factorise

```{r tidy_names}
raw_data %<>% 
  rename(ab_id_dichot = ABDERR_synth) %>% 
  mutate(ab_id_dichot = factor(ab_id_dichot,
                               levels = c(1, 2),
                               labels = c("Aboriginal ID",
                                          "Non-Aboriginal ID"))
         ) %>% 
  rename(ab_id_detailed = ABIDENT_synth) %>% 
  mutate(ab_id_detailed = factor(ab_id_detailed,
                                 levels = c(1:6),
                                 labels = c("North_Am_Indian", "Mètis",
                                            "Inuit",
                                            "Multiple Aborginal ID", 
                                            "Other Aboriginal",
                                            "Non-Aboriginal"))
         ) %>% 
  rename(adl_difficulty = ADIFCLTY_synth) %>% 
  mutate(adl_difficulty = factor(adl_difficulty,
                                 levels = c(1:4),
                                 labels = c("No", "Not stated", "Often",
                                            "Sometimes"))
         ) %>% 
  rename(age_imm = AGE_IMM_R_group_synth) %>% 
  mutate(age_imm = factor(age_imm,
                          levels = c(1:15),
                          labels = c("<5", "5 to <10", "10 to <15",
                                     "15 to <20", "20 to <25", "25 to <30",
                                     "30 to <35", "35 to <40", "40 to <45",
                                     "45 to <50", "50 to <55", "55 to <60",
                                     "60 and over", 
                                     "Non-permanent resident",
                                     "Non-imigrant and others"
                                     ))
         ) %>% 
  rename(citizen_stat = CITSM_synth) %>% 
  mutate(citizen_stat = factor(citizen_stat,
                               levels = c(1, 2),
                               labels = c("Canadian", "non-Canadian"))
         ) %>% 
  rename(cause_death_1 = COD1_synth) %>% 
  mutate(cause_death_1 = factor(cause_death_1,
                                levels = c(1:5),
                                labels = c("Communicable etc.", 
                                           "Noncommunicable", "Injuries",
                                           "Other causes or NA",
                                           "Did not Die"))
         ) %>% 
  rename(cause_death_2 = COD2_synth) %>% 
  mutate(cause_death_2 = factor(cause_death_2,
                                levels = c(1:14),
                                labels = c("Infectious diseases", 
                                           "Respiratory infections",
                                           "Colon and rectal cancers",
                                           "Breast cancers", "Diabetes",
                                           "Dementia", "IHD", "CVD",
                                           "Respiratory diseases",
                                           "Digestive diseases",
                                           "Genitourinary diseases",
                                           "Unintentional injuries",
                                           "Other causes or NA",
                                           "Did not Die"
                                           ))
         ) %>% 
  rename(worker_class = COWD_synth) %>% 
  mutate(worker_class = factor(worker_class,
                               levels = c(1:7),
                               labels = c("Unpaid family workers", 
                                          "Paid worker,no help",
                                          "Paid worker,paid help",
                                          "Paid Worker",
                                          "Self-employed,no help",
                                          "Self-employed,paid help",
                                          "Not applicable "))
         ) %>% 
  rename(adl_disab_difficulty = DISABFL_synth) %>% 
  mutate(adl_disab_difficulty = factor(adl_disab_difficulty,
                                       levels = c(1:4),
                                       labels = c("No", "Not stated",
                                                  "Often", "Sometimes"))
         ) %>% 
  rename(adl_disab_diff_type = DISABIL_synth) %>% 
  mutate(adl_disab_diff_type = factor(adl_disab_diff_type,
                                      levels = c(1:17),
                                      labels = c("ADL & Home & other",
                                                 "ADL & Home & Work/School",
                                                 "ADL & Home & Work/School & other",
                                                 "ADL & Work & other", "ADL & Home",
                                                 "ADL & other", "ADL & Work/school",
                                                 "ADL", "None", "Home & other",
                                                 "Home & Work/school",
                                                 "Home & Work/school & other",
                                                 "Work/school & other", "Home",
                                                 "Other", "Work/school",
                                                 "Not stated"))
         ) %>% 
  rename(birth_country = DPOB11N_synth) %>% 
  mutate(birth_country = factor(birth_country,
                                levels = c(1:10),
                                labels = c("non-Immigrant",
                                           "Latin America & Caribbean",
                                           "Western Europe", "Eastern Europe",
                                           "Sub-Saharan Africa",
                                           "N Africa, SW Asia & Middle East",
                                           "S Asia", "SE Asia", "E Asia", 
                                           "Australia, NZ, Oceania & Greenland"))
         ) %>% 
  rename(vis_minority = DVISMIN_synth) %>% 
  mutate(vis_minority = factor(vis_minority,
                               levels = c(1:14),
                               labels = c("Chinese", "South Asian", "Black",
                                          "Filipino", "Latin American", 
                                          "Southeast Asian", "Arab", "West Asian",
                                          "Korean", "Japanese", 
                                          "Other visible minority",
                                          "Multiple visible minority", 
                                          "Aboriginal self-reporting",
                                          "Not a visible minority"))
         ) %>% 
  rename(no_fam_members = EFCNT_PP_R_synth) %>% 
  mutate(no_fam_members = factor(no_fam_members,
                                 levels = c(1:10),
                                 labels = c("1", "2", "3", "4", "5",
                                            "6", "7", "8", "9", "10 or more"))
         ) %>% 
  rename(first_lang = FOL_synth) %>% 
  mutate(first_lang = factor(first_lang,
                             levels = c(1:4),
                             labels = c("English", "French", "English & French",
                                        "Other"))
         ) %>% 
  rename(employ_status = FPTIM_synth) %>% 
  mutate(employ_status = factor(employ_status,
                                levels = c(1:3),
                                labels = c("Full-time", "Part-time", "Didn't work"))
         ) %>% 
  #Note the definitions in the metadata file - 3rd generation does not mean
  #what I would intuitively think
  rename(generation = GENSTPOB_synth) %>% 
  mutate(generation = factor(generation,
                             levels = c(1:3),
                             labels = c("1st Generation", "2nd Generation",
                                        "3rd Generation"))
         ) %>% 
  rename(education = HCDD_synth) %>% 
  mutate(education = factor(education,
                            levels = c(1:13),
                            labels = c("None", "High School", "Trades Cert/Dip",
                                       "Apprenticeship Cert/Dip",
                                       "Non-Uni Cert/Dip < 1 year",
                                       "Non-Uni Cert/Dip 1-2 years",
                                       "Non-Uni Cert/Dip > 2years",
                                       "Uni Cert/Dip below Bachelor's",
                                       "Bachelor's degree",
                                       "Uni Cert/Dip above Bachelor's",
                                       "Medicine etc.", "Master's degree",
                                       "Doctorate"))
         ) %>% 
  rename(immigration_stat = IMMDER_synth) %>% 
  mutate(immigration_stat = factor(immigration_stat,
                                   levels = c(1:3),
                                   labels = c("Immigrant", "Non-permanent resident",
                                              "Non-immigrant"))
         ) %>% 
  rename(no_kids = KID_group_synth) %>% 
  mutate(no_kids = factor(no_kids,
                          levels = c(1:3),
                          labels = c("None", "1 or 2",
                                     "3 or more"))
         ) %>% 
  rename(lo_inc_aftertax = LOINCA_synth) %>% 
  mutate(lo_inc_aftertax = factor(lo_inc_aftertax,
                                  levels = c(1:3),
                                  labels = c("non-low income", "low income",
                                             "Concept not applicable"))
         ) %>% 
  rename(lo_inc_beforetax = LOINCB_synth) %>% 
  mutate(lo_inc_beforetax = factor(lo_inc_beforetax,
                                   levels = c(1:3),
                                  labels = c("non-low income", "low income",
                                             "Concept not applicable"))
         ) %>% 
  rename(mar_stat = MARST_synth) %>% 
  mutate(mar_stat = factor(mar_stat,
                           levels = c(1:5),
                           labels = c("Divorced", "Married", "Separated", 
                                      "Never married", "Widowed"))
         ) %>% 
  rename(occupation = NOCSBRD_synth) %>% 
  mutate(occupation = factor(occupation,
                             levels = c(1:11),
                             labels = c("Management", "Business", 
                                        "Science", "Health", "Govt & Religion",
                                        "Art & Culture", "Sales & Service",
                                        "Trades & Transport", "Primary industry",
                                        "Manufacturing", "Not Applicable"))
         ) %>% 
  rename(official_lang = OLN_synth) %>% 
  mutate(official_lang = factor(official_lang,
                                levels = c(1:4),
                                labels = c("English", "French", "English & French",
                                           "Other"))
         ) %>% 
  rename(place_of_birth = POBDER_synth) %>% 
  mutate(place_of_birth = factor(place_of_birth,
                                 levels = c(1:3),
                                 labels = c("Home province","Other province",
                                            "Outside Canada"))
         ) %>% 
  rename(province = PR_synth) %>% 
  mutate(province = factor(province,
                           levels = c(10, 11, 12, 13, 24, 35,
                                      46, 47, 48, 59, 60, 61, 62),
                           labels = c("Newfoundland and Labrador", 
                                      "Prince Edward Island", 
                                      "Nova Scotia", "New Brunswick", "Quebec",
                                      "Ontario", "Manitoba", "Saskatchewan",
                                      "Alberta", "British Columbia", "Yukon", 
                                      "Northwest Territories", "Nunavut"))
         ) %>% 
  rename(repairs = RPAIR_synth) %>% 
  mutate(repairs = factor(repairs,
                          levels = c(1:4),
                          labels = c("No", "Minor", "Major", "Not applicable"))
         ) %>% 
  rename(rur_urb = RUINDFG_synth) %>% 
  mutate(rur_urb = factor(rur_urb,
                          levels = c(1, 2),
                          labels = c("Rural", "Urban"))
        ) %>% 
  rename(sex = SEX_synth) %>% 
  mutate(sex = factor(sex,
                      levels = c(1, 2),
                      labels = c("Female", "Male"))
         ) %>% 
  rename(dead = S_DEAD_synth) %>% 
  mutate(dead = factor(dead,
                       levels = c(1,2),
                       labels = c("Dead", "Not Dead"))
         ) %>% 
  rename(transport = TRMODE_synth) %>% 
  mutate(transport = factor(transport,
                            levels = c(1:9),
                            labels = c("Bicycle", "Driver", "Motorbike",
                                       "Other", "Passenger", "Taxi", 
                                       "Public Transport", "Walk", "Not applicable"))
         ) %>% 
  rename(year_imm = YRIM_group_synth) %>% 
  mutate(year_imm = factor(year_imm,
                           levels = c(1:6),
                           labels = c("2002 or later", "1997-2001", "1987-1996",
                                      "1986 or earlier", "Non-permanent resident",
                                      "Non-immigrant and others"))
         ) %>% 
  rename(age_grp = age_group_synth) %>% 
  mutate(age_grp = factor(age_grp,
                          levels = c(1:15),
                          labels = c("19-24", "25-29", "30-34", "35-39", "40-44",
                                     "45-49", "50-54", "55-59", "60-64", "65-69",
                                     "70-74", "75-79", "80-84", "85-89", "90 plus"))
         ) %>% 
  rename(loinc_decile = d_licoratio_da_bef_synth) %>% 
  mutate(loinc_decile = factor(loinc_decile,
                               levels = c(1:10),
                               labels = c("1 - lowest", "2", "3", "4", "5",
                                          "6", "7", "8", "9", "10 - highest")))
  

```


###Save

I'll save the cleaned dataset with a new name `clean_data` which can be loaded directly in other analysis files

```{r save, eval=FALSE}
clean_data <- raw_data
save(clean_data, file = here("assets/clean_data/clean_data.rds"))
rm(clean_data)
```


```{r tidy_1, echo=FALSE}
rm(raw_data)
```


##Shapefile

**also try https://www.weather.gov/gis/CanadianProvinces**

Quick look at the shapefile to make sure everything is in order

```{r shape}
glimpse(shapefile)
```

Need to change one variable name and drop some unneeded variables

```{r shape_clean}
shapefile %<>% 
  rename(province = NAME) %>% 
  select(province, geometry)
```


And a quick plot of the blank file to check it looks ok

```{r plot_shape, fig.width=9, fig.height=6}
shapefile %>% 
  ggplot() +
  geom_sf() +
  theme(line = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_line(colour = "transparent"))
```

###Save it

```{r save_shp, eval=FALSE}
save(shapefile, file = here("assets/clean_data/shapefile.rds"))
```


#Session info

```{r session_info, cache=FALSE}
devtools::session_info()
```
